---
title: "Data visualization about US"
author: "Adrian Cabrera, Martí Rivero, Gerard Garcia"
date: "November 29, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(maps)
library(mapproj)
setwd("~/ProjecteAdrianGerardMarti/")
demographics_raw <- read_csv("chsi_dataset/DEMOGRAPHICS.csv", na = c("-1111.1", "-1111", "-1", "-2222", "-2222.2", "-2"))
measures_of_birth_and_death_raw <- read_csv("chsi_dataset/MEASURESOFBIRTHANDDEATH.csv", na = c("-1111.1", "-1111", "-1", "-2222", "-2222.2", "-2"))
risk_factors_and_access_to_care_raw <- read_csv("chsi_dataset/RISKFACTORSANDACCESSTOCARE.csv", na = c("-1111.1", "-1111", "-1", "-2222", "-2222.2", "-2"))
leading_causes_of_death_raw <- read_csv("chsi_dataset/LEADINGCAUSESOFDEATH.csv", na = c("-1111.1", "-1111", "-1", "-2222", "-2222.2", "-2"))
vulnerable_pops_and_env_health_raw <- read_csv("chsi_dataset/VUNERABLEPOPSANDENVHEALTH.csv", na = c("-1111.1", "-1111", "-1", "-2222", "-2222.2", "-2"))
preventive_services_use_raw <- read_csv("chsi_dataset/PREVENTIVESERVICESUSE.csv", na = c("-1111.1", "-1111", "-1", "-2222", "-2222.2", "-2"))
count_nas <- function(vec) {
  return(sum(is.na(vec)))
}
```

Importing the dataset
-
### Explaining the dataset, the data and references
This dataset contains various indicators of health related to diseases, safe living conditions, deaths, births ...

The data is organized in states and the 3141 counties of the United States with more than 200 variables in total.

This dataset is not designed only for healthcare professionals but for anyone who wants to do research or improve their community
*References*

[Dataset official webpage](https://healthdata.gov/dataset/community-health-status-indicators-chsi-combat-obesity-heart-disease-and-cancer).

## The columns
#### DEMOGRAPHICS COLUMNS

| Name                | Data Type     | Description                       |  
|---------------------|---------------|-----------------------------------|          
| `State_FIPS_Code`   | Text          |Two-digit state identifier         |                               
| `County_FIPS_Code`  | Text          |Three-digit county identifier      |                
| `CHSI_County_Name`  | Text          |Name of county                     |             
| `CHSI_State_Name`   | Text         |Name of State or District of Columbia| 
| `Population_Size`   |Integer        |County data, population size|
| `Poverty`   |Decimal        |County data, individuals living below poverty level|
| `White`   |Decimal        |County data, White|
| `Black`   |Decimal        |County data, Black|
| `Native_American`|Decimal        |County data, American Indian|
| `Asian`|Decimal        |County data, Asian/Pacific Islander|
| `Hispanic`|Decimal        |County data,Hispanic origin|

## Number of rows 
```{r echo=FALSE}
nrow(demographics_raw)
```
## Number of NA 
```{r echo=FALSE}
count_nas(demographics_raw)
```
#### MEASURESOFBIRTHANDDEATH COLUMNS

| Name                | Data Type     | Description                       |  
|---------------------|---------------|-----------------------------------|          
| `State_FIPS_Code`   | Text          |Two-digit state identifier         |                               
| `County_FIPS_Code`  | Text          |Three-digit county identifier      |                
| `CHSI_County_Name`  | Text          |Name of county                     |             
| `CHSI_State_Name`   | Text         |Name of State or District of Columbia| 
| `Infant_Mortality`   | Decimal         |County data, infant mortality|
| `Lung_Cancer`   | Decimal         |County data, death measures, lung cancer|
| `Suicide`   | Decimal         |County data, death measures, suicide|
| `MOBD_Time_Span`   | Text         |Time period of reported data for measures of birth and death|

## Number of rows 
```{r echo=FALSE}
nrow(measures_of_birth_and_death_raw)
```
## Number of NA 
```{r echo=FALSE}
count_nas(measures_of_birth_and_death_raw)
```
#### VULNETABLEPOPSANDENVHEALTH COLUMNS

| Name                | Data Type     | Description                       |  
|---------------------|---------------|-----------------------------------|          
| `State_FIPS_Code`   | Text          |Two-digit state identifier         |                            
| `County_FIPS_Code`  | Text          |Three-digit county identifier      |                
| `CHSI_County_Name`  | Text          |Name of county                     |             
| `CHSI_State_Name`   | Text         |Name of State or District of Columbia| 
| `Unemployed`   | Text              |County data, unemployed|

## Number of rows 
```{r echo=FALSE}
nrow(vulnerable_pops_and_env_health_raw)
```
## Number of NA 
```{r echo=FALSE}
count_nas(vulnerable_pops_and_env_health_raw)
```
#### PREVENTIVESERVICESUSE COLUMNS

| Name                | Data Type     | Description                       |  
|---------------------|---------------|-----------------------------------|          
| `State_FIPS_Code`   | Text          |Two-digit state identifier         |                            
| `County_FIPS_Code`  | Text          |Three-digit county identifier      |                
| `CHSI_County_Name`  | Text          |Name of county                     |             
| `CHSI_State_Name`   | Text         |Name of State or District of Columbia| 
| `HepA_Rpt`   | Integer         |County data, Hepatitis A reported cases|
| `HepB_Rpt`   | Integer         |County data, Hepatitis B reported cases|

## Number of rows 
```{r echo=FALSE}
nrow(preventive_services_use_raw)
```
## Number of NA 
```{r echo=FALSE}
count_nas(preventive_services_use_raw)
```
#### RISKFACTORSANDACCESSTOCARE COLUMNS

| Name                | Data Type     | Description                       |  
|---------------------|---------------|-----------------------------------|          
| `State_FIPS_Code`   | Text          |Two-digit state identifier         |                            
| `County_FIPS_Code`  | Text          |Three-digit county identifier      |                
| `CHSI_County_Name`  | Text          |Name of county                     |             
| `CHSI_State_Name`   | Text         |Name of State or District of Columbia|
| `Obesity`   | Decimal         |County data,obesity|
| `Smoker`   | Decimal         |County data,smoker| 
| `Uninsured`   | Integer         |County data,uninsured individuals| 

## Number of rows 
```{r echo=FALSE}
nrow(risk_factors_and_access_to_care_raw)
```
## Number of NA 
```{r echo=FALSE}
count_nas(risk_factors_and_access_to_care_raw)
```
### N/A information
In the dataset we have found a report that explains how the NAs are indicated.
According to the table we have found multiple values that we converted to NA during import. We indicate these values in the following relationship:
(values - additional information)

* -9999 - For unenployed variable.
* -2222, -2222.2, -2 OR nda - No data available
* -1111, -1111.1, -1 OR nrf - No report

Tidying the dataset
-
No, the dataset was not cleant. We had to delete colums related to intervals, quantiles, percentiles that we did not require in our case. We used regular expressions patterns, selects and filters to tidy the data as needed. 

In some situations we had to convert the data to long format. We invested and found the function pivot_longer() but we could not make use of it since is still in the maturation phrase and it is not included in the tidyverse standard, only in the development version, then we find the function gather with which achieved that change.

The plots
-
Used libraries:

`library(tidyverse)`

`library(maps)`

`library(mapproj)`


Tidying data:

```{r echo=FALSE}
# Import data:
# We consider NA the following values:
# -111.1, -1111, -1, -2222, -2222.2, -2


# Vector of string of column names
demographics_colname <- colnames(demographics_raw)
measures_of_birth_and_death_colname <- colnames(measures_of_birth_and_death_raw)
risk_factors_and_access_to_care_colname <- colnames(risk_factors_and_access_to_care_raw)
leading_causes_of_death_colname <- colnames(leading_causes_of_death_raw)
vulnerable_pops_and_env_health_colname <- colnames(vulnerable_pops_and_env_health_raw)
preventive_services_use_colname <- colnames(preventive_services_use_raw)

# Create a pattern variable to delete all coulmns containing one of the following strings
# Ind, Max, Min, CI
pattern <- "Ind|Max|Min|CI"

# Filter the colnames using the pattern created
# We obtain an int vector
demographics_result <- grep(pattern, demographics_colname)
measures_of_birth_and_death_result <- grep(pattern, measures_of_birth_and_death_colname)
risk_factors_and_access_to_care_result <- grep(pattern, risk_factors_and_access_to_care_colname)
leading_causes_of_death_result <- grep(pattern, leading_causes_of_death_colname)
vulnerable_pops_and_env_health_result <- grep(pattern, vulnerable_pops_and_env_health_colname)
preventive_services_use_result <- grep(pattern, preventive_services_use_colname)

# Tidy all tables using the vector created
demographics_tidy <- select(demographics_raw, -demographics_result)
measures_of_birth_and_death_tidy <- select(measures_of_birth_and_death_raw, -measures_of_birth_and_death_result)
risk_factors_and_access_to_care_tidy <- select(risk_factors_and_access_to_care_raw, -risk_factors_and_access_to_care_result)
leading_causes_of_death_tidy <- select(leading_causes_of_death_raw, -leading_causes_of_death_result)
vulnerable_pops_and_env_health_tidy <- select(vulnerable_pops_and_env_health_raw, -vulnerable_pops_and_env_health_result)
preventive_services_use_tidy <- select(preventive_services_use_raw, -preventive_services_use_result)
```

#### Exercise 3
Plot which of the most populated states has the gratest infantil mortality.

Result: We can see that texas is the state with the highest growth.

```{r echo=FALSE}
# Sum of the infant mortality of each state
states_infant_mortality_tidy <- measures_of_birth_and_death_tidy %>%
  group_by(CHSI_State_Name, MOBD_Time_Span) %>%
  summarise(Infant_Mortality = sum(Infant_Mortality, na.rm = TRUE))

# Sum of the population of each state
population <- demographics_tidy %>%
  group_by(CHSI_State_Name) %>%
  summarise(population = sum(Population_Size))

# Get the most populated states
more_populated_state <- head(arrange(population, desc(population)))

# Get a vector of the most populated states
more_populated_state_name <- more_populated_state$CHSI_State_Name

# Filter the table with population and infant mortality using the previous vector
infant_mortality_in_state <- filter(states_infant_mortality_tidy, CHSI_State_Name %in% more_populated_state_name)

# Plot of the result
ggplot(infant_mortality_in_state, aes(x = CHSI_State_Name, y = Infant_Mortality)) + geom_col(fill = "#FFDB6D", color = "gray") + facet_grid(MOBD_Time_Span ~ .) +
  labs(
    title = "Infanitl mortality cases",
    subtitle = "Infanitl mortality cases in the six more populated states and divided by periods",
    x = "States",
    y = "Infant mortality cases",
    caption = "Last update 2003"
  )
```

#### Exercise 4
Plot the number of smokers and the lung cancer deaths in the six states with the highest number of smokers.

Result: We cannot arrive at indications of a causation given that for example the state of Elliot County has the lowest percentage of smokers and is the second with the most deaths from lung cancer.

```{r echo=FALSE}
# Filter the table to get the data from 1994 to 2003
measures_of_birth_and_death_tidy_1994_2003 <- filter(measures_of_birth_and_death_tidy, MOBD_Time_Span == "1994-2003")

# Drop NA from Smokers column
risk_factors_and_access_to_care_without_na <- risk_factors_and_access_to_care_tidy %>% drop_na(Smoker)

# Join tables to get the state name, county name, number of lung cancer and the percentage of smokers
lung_cancer_smokers_join_table <- inner_join(measures_of_birth_and_death_tidy_1994_2003, risk_factors_and_access_to_care_without_na, by = c("CHSI_County_Name", "CHSI_State_Name"))
lung_cancer_smokers_join_table <- lung_cancer_smokers_join_table %>%
  select(CHSI_State_Name, CHSI_County_Name, Lung_Cancer, Smoker) %>%
  arrange(desc(Smoker)) %>%
  head()

# Plot the result
ggplot(lung_cancer_smokers_join_table, aes(x = CHSI_County_Name)) +
  geom_point(aes(y = Lung_Cancer, color = "Lung cancer")) +
  geom_point(aes(y = Smoker, color = "Smokers")) +
  labs(
    title = "Lung cancer and Smokers",
    subtitle = "Relation between lung cancer and smokers in some states",
    x = "States",
    y = "Number",
    caption = "Last update 2003"
  ) +
  scale_color_manual(name = "Leyenda", values = c("red", "blue"))
```


#### Exercise 5
Plot the percentage of races that commit suicide in Bronx.

Result: We observe that the majority races are black and white.

```{r echo=FALSE}

# Filter the tabe to get all data from Bronx
bronx_data_raw <- demographics_tidy %>% filter(CHSI_County_Name == "Bronx")

# Get the columns of each race
bronx_data <- select(bronx_data_raw, White:Asian)

# There is a small gap, we filled that gap with Uknown
# We consider Uknow as undocumented people
bronx_tidy <- mutate(bronx_data, "Unknown" = 100 - sum(White + Black + Native_American + Asian))

# Create a tibble with the races and the values
result <- tibble(races = colnames(bronx_tidy), values = as.numeric(c(bronx_tidy[1, 1:ncol(bronx_tidy)])))

# Plot the result
ggplot(result, aes(x = "", y = values, fill = races)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  labs(
    title = "Percentage of each race that commit suicide in Bronx",
    subtitle = "Percentage of races in each State of United States",
    x = "States",
    y = "Percentage",
    fill = "Legend",
    caption = "Last update 2003"
  )
```

#### Exercise 6
Compare the percentage of each race that are in the east and west coast.

Result: We can see that on the east coast there is a higher percentage of black population than on the west coast.

```{r echo=FALSE, fig.width=16, fih.height=7}

# Set the color pattern to use it in the plot
cbbPalette <- c("#EFD36E", "#592f2a", "#E6A96F", "#FFDBAC")

# EAST

# Get the state names from the east coast
east_coast_state_name <- c("Maine", "New Hampshire", "Massachusetts", "Rhode Island", "Connecticut", "New York", "New Jersey", "Pennsylvania", "Delaware", "Maryland", "Virginia", "North Carolina", "South Carolina", "Georgia", "Florida")

# Filter the table to get all data of the east coast states
east_coast_state <- filter(demographics_tidy, CHSI_State_Name %in% east_coast_state_name)

# Select the state name and races columns of the east coast states
east_coast_states_race <- select(east_coast_state, CHSI_State_Name, White:Asian)

# Group by east coast states and summarise each race
east_coast_states_race_tidy <- east_coast_states_race %>%
  group_by(CHSI_State_Name) %>%
  summarise(White = mean(White), Black = mean(Black), Native_American = mean(Native_American), Asian = mean(Asian))

# Make the table to long format and group by state name
table_formateada_east_gather <- east_coast_states_race_tidy %>%
  gather(Total, Value, -CHSI_State_Name) %>%
  group_by(CHSI_State_Name)

# Plot the result
ggplot(table_formateada_east_gather, aes(x = CHSI_State_Name, y = Value, fill = Total)) +
  geom_col(position = "dodge") +
  labs(
    title = "Percentage of races in East Coast",
    subtitle = "Percentage of races in each State of United States",
    x = "States",
    y = "Percentage",
    fill = "Legend",
    caption = "Last update 2003"
  ) +
  scale_fill_manual(values = c(cbbPalette))

# WEST

# Get the state names from the west coast
west_coast_state_name <- c("California", "Nevada", "Arizona", "Orlando", "Washington", "Alaska")

# Filter the table to get all data of the west coast states
states_west_coast_tbl <- filter(demographics_tidy, CHSI_State_Name %in% west_coast_state_name)

# Select the state name and races columns of the west coast states
west_coast_states_race <- select(states_west_coast_tbl, CHSI_State_Name, White:Asian)

# Group by west coast states and summarise each race
west_coast_states_race_tidy <- west_coast_states_race %>% group_by(CHSI_State_Name) %>% summarise(White = mean(White), Black = mean(Black), Native_American = mean(Native_American), Asian = mean(Asian))

# Make the table to long format and group by state name
table_formateada_west_gather <- west_coast_states_race_tidy %>%
  gather(Total, Value, -CHSI_State_Name) %>%
  group_by(CHSI_State_Name)

# Plot the result
ggplot(table_formateada_west_gather, aes(x = CHSI_State_Name, y = Value, fill = Total)) +
  geom_col(position = "dodge") +
  labs(
    title = "Percentage of races in West Coast",
    subtitle = "Percentage of races in each State of United States",
    x = "States",
    y = "Percentage",
    fill = "Legend",
    caption = "Last update 2003"
  ) +
  scale_fill_manual(values = c(cbbPalette))
```

#### Exercise 7
Plot the states with most poverty and compare to its obesity rates.

Result: We have no evidence of causality as obesity is fairly uniform across states while poverty is much more variable.

```{r echo=FALSE}
# Group by state names and then create a mean of obesity for each state
states_obesity <- risk_factors_and_access_to_care_tidy %>%
  group_by(CHSI_State_Name) %>%
  summarise(obesity = mean(Obesity, na.rm = T))

# Group by state names and then create a mean of poverty for each state
states_population_size <- demographics_tidy %>%
  group_by(CHSI_State_Name) %>%
  summarise(poverty = mean(Poverty, na.rm = T))

# Join  the poverty and the obesity means
join_tables <- inner_join(states_obesity, states_population_size, by = c("CHSI_State_Name", "CHSI_State_Name"))

# Get the states with more obesity and poverty
major_ordered_by_obesity_poverty <- head(arrange(join_tables, desc(poverty), desc(obesity)))

# Make the table long format
longer_format <- major_ordered_by_obesity_poverty %>%
  gather(Total, Value, -CHSI_State_Name)

# Plot the result
ggplot(longer_format, aes(x = CHSI_State_Name, y = Value, fill = Total)) +
  geom_col(position = "dodge") +
  labs(
    title = "Poverty and obesity in the highest poverty states",
    subtitle = "Poverty and obesity",
    x = "States",
    y = "Percentage",
    fill = "Legend",
    caption = "Last update 2003"
  )
```

#### Exercise 8
How  many counties exceed 20% of poverty and what are their names.

Result: We can see that the southern states are generally poorer than the northern ones.

```{r echo=FALSE}

# Get the data from the USA map (longitude and latitude of each state)
us_state_map <- map_data("state")

# Select the state name and poverty of each state
states_exceed_20_raw <- select(demographics_tidy, CHSI_State_Name, Poverty)

# Order the data in descending order by poverty
states_exceed_20_arrange <- arrange(states_exceed_20_raw, desc(Poverty))

# Group by state name and create a mean of poverty removin all NA
states_exceed_20_arrange <- states_exceed_20_arrange %>%
  group_by(CHSI_State_Name) %>%
  summarise(Poverty = mean(Poverty, na.rm = TRUE))

# Convert to lower case the state names to make the join
states_exceed_20_arrange$CHSI_State_Name <- tolower(states_exceed_20_arrange$CHSI_State_Name)

# Join both tables to get the longitude, latitude, poverty and state name of each state
data_usa <- merge(us_state_map, states_exceed_20_arrange, by.x = "region", by.y = "CHSI_State_Name")

# PLot the result
ggplot(data_usa, aes(x = long, y = lat, group = group, fill = Poverty)) +
  geom_polygon(color = "gray") +
  coord_map("polyconic") +
  scale_fill_gradient2() +
  labs(
    title = "Poverty percentage",
    subtitle = "Poverty percentage in each state of United States",
    x = "Longitude",
    y = "Latitude",
    fill = "Legend",
    caption = "Last update 2003"
  )
```

#### Exercise 9
Compare the richest and poorest counties by obesity and unemployment.

Result: We cannot draw any conclusions given the lack of data.

```{r echo=FALSE}

# join tables
join_tables_exc9 <- inner_join(risk_factors_and_access_to_care_tidy, demographics_tidy, by = "CHSI_County_Name") %>% inner_join(., vulnerable_pops_and_env_health_tidy, by = "CHSI_County_Name")

# select columns
most_poverty_state <- select(join_tables_exc9, CHSI_County_Name, Population_Size, Obesity, Poverty, Unemployed)

# select most poverty states
# look the most poverty states with arrange
arrange(most_poverty_state, desc(Poverty))

# save in c() names
name_most_poverty <- c("Starr", "Shannon", "East Carroll", "Todd")
# filter most poorest 4 states
most_poverty_state <- most_poverty_state %>% filter(CHSI_County_Name %in% name_most_poverty)

# group by  and convert table to format large
table_format_most_poverty_state <- most_poverty_state %>%
  group_by(CHSI_County_Name) %>%
  summarise(Poverty = mean(Poverty), Obesity = mean(Obesity), Unemployed = mean(Unemployed / Population_Size) * 100) %>%
  gather(Condition, Value, -CHSI_County_Name)
# graphic
ggplot(table_format_most_poverty_state, aes(x = CHSI_County_Name, y = Value, fill = Condition)) +
  geom_col(position = "dodge") +
  labs(
    title = "The poorest counties",
    subtitle = "Relationship of poverty, obesity and unemployment",
    x = "States",
    y = "Percentage",
    fill = "Condition",
    caption = "Last update 2003"
  )


# arrange
most_rich_state <- select(join_tables_exc9, CHSI_County_Name, Population_Size, Obesity, Poverty, Unemployed)
# select 6 rich states
# look the most poverty states with arrange
arrange(most_rich_state, Poverty)
# save in c() names
name_most_rich <- c("Douglas", "Los Alamos", "Falls Church City", "Hunterdon")
# filter most poorest 4 states
most_rich_state <- most_rich_state %>% filter(CHSI_County_Name %in% name_most_rich)
# group by  and convert table to format large
table_format_most_rich_state <- most_rich_state %>%
  group_by(CHSI_County_Name) %>%
  summarise(Poverty = mean(Poverty), Obesity = mean(Obesity), Unemployed = mean(Unemployed / Population_Size) * 100) %>%
  gather(Condition, Value, -CHSI_County_Name)

ggplot(table_format_most_rich_state, aes(x = CHSI_County_Name, y = Value, fill = Condition)) +
  geom_col(position = "dodge") +
  labs(
    title = "The richest states",
    subtitle = "Relationship of poverty, obesity and unemployment",
    x = "States",
    y = "Percentage",
    fill = "Condition",
    caption = "Last update 2003"
  )
```

#### Exercice 10

Compare the number of patients with Hepatitis A and B with the level of the population without health insurance.

Result: Causation cannot be established because a county like Fulton with very few insured has practically the same number of Hepatitis A and B patients.

```{r echo=FALSE}

# Adding up Hepatitis A and B values in new column. Ascendant ordering and getting the six first values.
hepatits_a_and_b <- select(preventive_services_use_tidy, CHSI_County_Name, CHSI_State_Name, HepA_Rpt, HepB_Rpt) %>%
  mutate(Hep_Total = HepA_Rpt + HepB_Rpt) %>%
  select(CHSI_County_Name, CHSI_State_Name, Hep_Total) %>%
  arrange(desc(Hep_Total)) %>%
  head()

# Selecting state info and insured people number.
uninsured <- select(risk_factors_and_access_to_care_tidy, CHSI_State_Name, CHSI_County_Name, Uninsured)

# Joined uninsured and hepatitis in one tibble. Then converted to long format.
hepatitis_and_uninsured <- inner_join(uninsured, hepatits_a_and_b, by = c("CHSI_State_Name", "CHSI_County_Name")) %>%
  select(-CHSI_State_Name) %>%
  gather(Type, Value, -CHSI_County_Name)

# Plotting hepatitis cases and uninsured people by the selected states in two different tables for each variable
ggplot(hepatitis_and_uninsured, aes(x = CHSI_County_Name, y = Value, fill = Type)) +
  geom_col(position = "dodge") + facet_grid(Type ~ ., scales = "free") +
  labs(
    title = "Uninsured individuals and Hepatitis cases in some counties",
    subtitle = "Unisured and Hepatitis A and B",
    x = "States",
    y = "Number of cases",
    fill = "Legend",
    caption = "Last update 2003"
  )
  ```

